---
title: "Project 2: Penguin Data Analysis"
author: "Gerardo Gonzalez"
date: "April 2025"
format: html
output-dir: docs
---

# Setup

This loads the R script so we can include the code chunks below.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk("Project-2-Final-Cleaned-script.R")
```

# Welcome!

Welcome to Project 2 of my Palmer Penguin data analysis for ZOOL710 at the University of Hawai'i at Manoa with Dr. Marguerite Butler.

In this project, we focus on statistical analysis using the cleaned dataset from Project 1. Here, I explore variation in penguin body mass and how it differs across species and islands.

These are the questions I aim to answer:

1. Do penguin species differ significantly in body mass?
2. Do both species and island influence penguin body mass?
3. How does body mass vary across species and islands?

Let's walk through Project 2 together!

# Load Packages

Let's first load the packages we need.

```{r, Packages, message = FALSE, warning = FALSE}
```

# Load Data

Now let's load the cleaned dataset we created in Project 1.

```{r, LoadData}
```

# Summary Table

We’ll use `skimr` to generate a detailed summary of our dataset.

```{r, SummaryTable}
```

# Count total number of observations in the dataset

```{r, TotalObservations}
```

# Calculate N per variable (non-missing values for each column)

```{r, Nper}
```

# Select and rename relevant columns from the skim summary

```{r, Relcol}
```

# Compute standard error for continuous variables

```{r, Standarderror}
```

# Print out the cleaned and renamed summary table

```{r, Viewtable}
```

# Question 1: Do Species Differ in Body Mass?

Let's test if different penguin species have significantly different body masses.

### Boxplot of Body Mass by Species

We start with a visual comparison using a boxplot.

```{r, BoxplotMassBySpecies}
```

### ANOVA: Body Mass ~ Species

Now we run an ANOVA to test the statistical significance of the differences.

```{r, ANOVA1}
```

# Question 2: Do Species and Island Affect Body Mass?

We extend our analysis to see if Island plays a role in explaining mass variation.

### ANOVA: Body Mass ~ Species + Island

```{r, ANOVA2}
```

# Visualize the Interaction Between Species and Island

Let's create a grouped boxplot to explore how species and island interact in relation to body mass.

```{r, InteractionPlot}
```

# Question 3: Body Mass Distribution by Species and Island

We use a faceted density plot to visualize the distribution of body mass by species on each island.

```{r, DensityPlot}
```

# Save Summary Table (Optional)

Finally, we save the summary table we created in case we want to use it in another analysis or report.

```{r, SaveClean}
```

# All Done!

Thanks for following along in Project 2. This completes my analysis of penguin body mass variation using statistical modeling and data visualization!
