---
title: "Project 1: Penguin Data Cleaning and Analysis"
author: "Gerardo Gonzalez"
date: "March 27, 2025"
output: html_document
---

# Setup

This needs to run to load the R script so we can include the code chunks below.
Note that you can pull in more than one R script if you want.

```{r, include=FALSE, cache=FALSE}
knitr::read_chunk("Project-1-Final-Cleaned-script.R")
```

# Welcome!

Hello and welcome to my Palmer Penguin Data Analysis project from Dr. Marguerite Butler for the ZOOL710 class here at the University of Hawai'i at Manoa.

In Project 1, the main focus is to:

	1) Clean the palmer penguins data set
	2) Analyze it further
	3) Make it reproducible for others to use.

In this document I walk you through how i corrected the errors in the data set as well as all my commands with comments.

Let's walk through project 1 together!



# Load Required Packages 

```{r, Packages, message = FALSE, warning = FALSE}
```

# Load Data

After you copy the repository and get in on your computer, Navigate to the Processing Code folder. 

Now let's Load it up!

```{r, Loaddata}
```

# Explore Data

Now we can explore the data and first get a glimpse of what it looks like.
We can do this by using the glimpse, summary, and skimr functions.
Look through the data and see of there are any errors or anything unusual we may need to fix.

```{r, Exploredata}
```


# Error 1: Fix Species Typos

This dataset should only have 3 species, but we actually have 9.
Let's look at the different species using the "unique" command.

```{r, Cleandata1}
```
Here we can easily see all the typos, lets fix them.

First we'll make a working copy of the raw data so we can compare our modified version to the original later on.

```{r, Createcopy}
```

Now we can fix the species names.

```{r, Cleandata2}
```

Looks Good!



# Error 2: Missing Values

When you use the dplyr::glimpse(rawdata) command, we notice there is a missing value for Culmen length.
We need to fix this by converting "missing" values to NA and ensure numeric type for Culmen Length.

```{r, Fixmissing}
```

Looks Good!



# Error 3: Irregular Data Points

We can use plots to look at the distribution of the data just to get a visual and see of anything looks off.
We can plot a histogram of culmen length and frequency, and a plot of body mass and culmen length.

```{r, Idirregulars}
```

Fix the Outliers 

In the histogram, 3 penguins have extremely long culmen lengths (300+mm) which is most likely a typo that we need to remove.

```{r, Fixbeak}
```

Looks Good! 



# Error 4: Remove Incorrect Body Mass Values 

In the Bivariate plot, we can see some extremely light penguins. These are most likely babies and aren't relevant to the data set.

The data focuses on adults. Let's go ahead and remove them.

```{r, Bodymass}
```

Looks Good!


# Error 5: Missing Masses

Mass is the main size variable, so we need to remove individuals with misssing masses in order to analyze the data. 

```{r, Missingmass}
```

Looks Good!


# Have Species, Sex, and Island coded as categorical/factor variable

```{r, Category}
```

Looks Good! 



# Extra Analysis: Make Bivariate Plots

We can make some bivariate plots for any remaining continuous data to ensure there are no further errors.
We can make many different plots and just have fun exploring the visual representations of the relationships between different variables. 

### Scatter plot 1: Body Mass vs. Culmen Depth

This plot looks at if the heavier penguins tend to have deeper culmens (beaks).

A positive relationship could mean that larger penguins have proportionally larger beak depths.

```{r, Scatter1}
```


### Scatter plot 2: Culmen Length vs. Culmen Depth

This graph explores the relationship between beak length and beak depth.

This could reveal adaptations to diet or feeding methods.

```{r, Scatter2}
```

### Scatter plot 3: Flipper Length vs. Body Mass

This plot looks at of penguins with longer flippers tend to have greater body mass.

Penguins with longer flipers may correlate to better swimming efficieny and reflect overall body condition.

```{r, Scatter3}
```

### Boxplot 1: Body Mass by Species

This box plot compares the body mass distribution across the three penguin species.

```{r, Box1}
```


### Boxplot 2: Culmen Length by Island

This boxplot looks at the difference in beak length among penguins from different islands.

Variations might reveal island specific ecological or genetic influences on penguin morphology

```{r, Box2}
```


### Density plot 1: Body Mass by Species

This plot looks at the body mass distribution for each penguin species.

```{r, Dens1}
```


### Density plot 2: Flipper Length by Sex

This plot looks at the differences in flipper length between males and females and if there are any major differences.

```{r, Dens2}
```


### Histogram of Body Mass by Species

This histogram is just another way to show the distribution of body mass by species.

```{r, Histogram}
```


### Bar plot: Penguin Count per Island

This bar plot summarizes the number of penguins sampled from each island.

This can help identify possible sampling biases ot differens in penguin population distribution across islands.

```{r, Barplot}
```


### Violin plot: Flipper Length by Species

This plot shows how flipper length varies within and between species.

The width indicates the frequency of penguins at a given flipper length.

```{r, Violin}
```


### Faceted scatter plot

This plot breaks down the relationship between body mass and culmen length separately by species.

It shows if the relationship differs among species and helps identify species-specific patterns or outliers.

```{r, Faceted}
```



# Finalize the Data

Final Clean 

```{r, Finalclean}
```

Final Check

```{r, Finalcheck}
```



# Save the Data

Save processed data as an R file and csv in a new folder

For me it will say it already exists because I made a folder for it already, so don't mind that message.
For you it should work.

```{r, Finalsave}
```


# All Done...Great Job!

